Pool Model - A Mathematical Framework\\ for Modelling Bacterial Consortia in Food Science

### 2025-09-26 paper/pool_model.tex
correct diffusion constant
migrate to only xcolor
add two more sentences for description of figure
Corrections from Christian

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[
    backend=biber,
    sorting=none,
    style=numeric-comp,
    hyperref=true,
    maxnames=2,
    url=false,
    isbn=false,
    eprint=false,
    date=year,
    doi=false,
    giveninits=true
]{biblatex}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{color}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
\usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

\title{Pool Model - A unified Mathematical\\ Modelling Framework for Food Science}
\author{
    Polina Gaindrik\textsuperscript{1},
    Jonas Pleyer\textsuperscript{1}
    and Christian Fleck\textsuperscript{1}
}

\date{
    \footnotesize
    \textsuperscript{\textbf{1}}Freiburg Center for Data Analysis and Modeling
}

@@

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from interactions that lead to
    inter-species cooperation or competition.
    Various empirical and inactivation models have been conceived and applied to problems related to
    food microbiology.
    However, the utilized models live on the popluation level, omitting any spatial effects such as
    heterogeneity or diffusion.
    In this work we present a novel generalized framework, called "pool model" which uses multiple
    distinct pools that can hold bacterial states of distinct species or extracellular nutrient
    abundance.
    It allows us to describe these systems mechanistically as variations of an underlying, more
    fundamental model.
    To illustrate the usefulness of the model, multiple different examples are reported.
    Finally, we construct an \acl{abm} for a given pool model, showcasing that the resulting model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of spatial effects on the dynamcis of bacterial
    growth.
    % \textbf{Notes from Christian}
    % Different empirical and inactivation models applied in the estimation
    % of parameters and modelling of phenomena in food microbiology, gen-
    % eralized Lotka Voltera approach used to describe the interaction of mi-
    % croorganisms, mechanistic models utilized to provide more insight into
    % the underlying dynamic can all be unified as variations of a new model
    % known as three pool model. Compared against the aforementioned
    % methods which are applicable in specific applications with some limi-
    % tations, the present model turns out to be less complex with general
    % applicability, low number of parameters and preferably to explain the
    % mechanism of the phenomenon of interest. To illustrate the generality
    % of the model, three different examples are reported.
\end{abstract}

@@

    G &\stackrel{\mu'}{\longrightarrow} & D\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset

@@

    \dot{D} &= \mu  L + \mu' G- \beta D  

@@

    G + R_0  \stackrel{\alpha_0}{\longrightarrow} 2G.

@@

Note that the pools $L$, $G$ and $D$ represent the number of bacteria (or number of bacteria per unit area/volume) while $R_0$ represents an abstract resource pool.

@@

Then $R_0=N_R/V$ and $[R_0]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol).

@@

For simplicity, we make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and dimensionless resource pool $R(t) = R_0 / N_t$ ($0 \leqslant R \leqslant 1$).

@@

In this case the growth term can be simply written as $\alpha G R$ and the model is fully formulated in the following way:

@@

        \dot{G} &= \lambda L + \alpha R G-\mu' G\\
        \dot{D} &= \mu  L + \mu' G- \beta D\\
        \dot{R} &= - \frac{\alpha}{N_t} R G

@@

Such a model can be simplified by exploiting that any limitation of the resource pool also leads to a limit of the total bacterial count

@@

    R = 1 - \frac{L+G+D}{N_t}

@@

        The time dependence of the total bacterial count $N(t) = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:ode_3pool_resource2}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.

@@

        Other curves show the case of no bacteria dying $\mu=0$ and represent different lengths of lag phases: short $\lambda=10^{-2}$ (dark green), intermediate $\lambda=10^{-3}$ (light green) and long one $\lambda=10^{-4}$ (yellow).

@@

The dynamics of the three pool model is captured by equations~\ref{eq:ode_3pool_resource2} which can be used to derive the two pool model by setting $D\equiv 0$ and replacing $\mu$ with $\beta$).
For $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:

@@

        D(t) &= N_0 \mu e^{-\beta t}\frac{1-e^{-(\mu+\lambda-\beta)t}}{\mu+\lambda-\beta}

@@

For constant parameters $\mu$, $\lambda$, and $\beta$ the integrals for $L$ and $D$ can be inserted into the differential equation for $G$.

@@

        \dot{N} &= - \beta D + \alpha G\left(1-\frac{N}{N_t}\right)

@@

For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = \lim_{t\to\infty} D(t) = 0$.
It follows immediately that $\lim_{t\to\infty} G(t) = N_t$.
The equation exhibits a lag phase and an initial drop due to cell death.
To find a substantial initial drop one needs $\beta \mu \gg \alpha \lambda$.
The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.

@@

    1&= 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\

@@

    G &\stackrel{\gamma(t)}{\longrightarrow} & L\\
    D &\stackrel{\beta}{\longrightarrow} & \emptyset

@@

        \dot{D} &= \mu  L - \beta D

@@

@@

        \subcaption{Rand. $D=5.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Random initial cell distribution, $D_R=D_I=5.0$\unit{\micro\metre\squared\per\minute}}

>>>>

\setlength{\tabcolsep}{1.3pt}
\setlength{\columnsep}{0.4cm}
\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[
    backend=biber,
    sorting=none,
    style=numeric-comp,
    hyperref=true,
    maxnames=2,
    url=false,
    isbn=false,
    eprint=false,
    date=year,
    doi=false,
    giveninits=true
]{biblatex}
\usepackage{array}
\usepackage{bm}
\usepackage{booktabs}
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off, font=small]{caption}
\usepackage{diagbox}
\usepackage{float}
\usepackage[T1]{fontenc}
\usepackage[a4paper,inner=2cm,outer=2cm,top=2.5cm,bottom=2.5cm]{geometry}
\usepackage{graphicx}% Include figure files
\usepackage{here}
\usepackage{hyperref}
\usepackage{lipsum}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{mathtools}
\usepackage{mathrsfs}
\usepackage{multirow}
\usepackage{newfloat}
\usepackage{onimage}
\usepackage{siunitx}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{tikz}
\usepackage[dvipsnames]{xcolor}
\usepackage{xpatch}

@@

\title{Pool Model - A Mathematical Framework\\ for Modelling Bacterial Consortia in Food Science}
\author{
    Polina Gaindrik\textsuperscript{1},
    Jonas Pleyer\textsuperscript{1}
    and Christian Fleck\textsuperscript{1}
}

\date{
    \footnotesize
    \textsuperscript{\textbf{1}}Freiburg Center for Data Analysis and Modeling and AI
}

@@

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from interactions that lead to
    inter-species cooperation or competition.
    Various empirical and inactivation models have been conceived and applied to problems related to
    food microbiology.
    However, the utilized models live on the population level, omitting any spatial effects such as
    heterogeneity or diffusion.
    In this work we present a novel generalized framework, called "pool model" which uses multiple
    distinct pools that can hold bacterial states of distinct species or extracellular nutrient
    abundance.
    It allows us to describe these systems mechanistically as variations of an underlying, more
    fundamental model.
    To illustrate the usefulness of the model, multiple different examples are reported.
    Finally, we construct an \acl{abm} for a given pool model, showcasing that the resulting model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of spatial effects on the dynamcis of bacterial
    growth.
    % \textbf{Notes from Christian}
    % Different empirical and inactivation models applied in the estimation
    % of parameters and modelling of phenomena in food microbiology, gen-
    % eralized Lotka Voltera approach used to describe the interaction of mi-
    % croorganisms, mechanistic models utilized to provide more insight into
    % the underlying dynamic can all be unified as variations of a new model
    % known as three pool model. Compared against the aforementioned
    % methods which are applicable in specific applications with some limi-
    % tations, the present model turns out to be less complex with general
    % applicability, low number of parameters and preferably to explain the
    % mechanism of the phenomenon of interest. To illustrate the generality
    % of the model, three different examples are reported.
\end{abstract}

@@

    G &\stackrel{\mu'}{\longrightarrow} & D

@@

    \dot{D} &= \mu  L + \mu' G 

@@

    G + R  \stackrel{\alpha_0}{\longrightarrow} 2G.

@@

Note that the pools $L$, $G$ and $D$ represent the number of bacteria (or number of bacteria per unit area/volume) while $R$ represents an abstract resource pool.

@@

Then $R=N_R/V$ and $[R]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol).

@@

For simplicity, we make a transition to the dimensionless constant $\alpha:=\alpha_0 N_t$ and dimensionless resource pool $\tilde{R}(t) = R/ N_t$ ($0 \leqslant \tilde{R} \leqslant 1$).

@@

In this case the growth term can be simply written as $\alpha G \tilde{R}$ and the model is fully formulated in the following way:

@@

        \dot{G} &= \lambda L + \alpha \tilde{R} G-\mu' G\\
        \dot{D} &= \mu  L + \mu' G\\
        \dot{\tilde{R}} &= - \frac{\alpha}{N_t} \tilde{R} G

@@

This can be simplified by exploiting the conservation of the resource pool:

@@

    \tilde{R} = 1 - \frac{L+G+D}{N_t}

@@

        The time dependence of the bacterial count $N(t) = L(t) + G(t) $ is shown.
        All curves were calculated using the model (\ref{eq:ode_3pool_resource2}) with parameters $\alpha=2.5$, $\mu'=0.5$.

@@

        Other curves show the case of no bacteria dying $\mu=0$ and represent different lengths of lag phases: short $\lambda=10^{-2}$ (dark green), intermediate $\lambda=10^{-3}$ (light green) and long one $\lambda=10^{-4}$ (yellow). Note that the steady state for $N(t)$ and $D(t)$ is for this model $\lim_{t\to\infty} N(t)=0$ and $\lim_{t\to\infty} D(t)=N_t$, because of the depletion of the resource pool. 

@@

The dynamics of the three pool model is captured by equations~\ref{eq:ode_3pool_resource2} which can be used to derive the two pool model simply by omitting $D$.
Moreover, for $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:

@@

        D(t) &= N_0 \mu \frac{1-e^{-(\mu+\lambda)t}}{\mu+\lambda

@@

For constant parameters $\mu$ and $\lambda$ the integrals for $L$ and $D$ can be inserted into the differential equation for $G$.

@@

        \dot{N} &= \alpha G\left(1-\frac{N}{N_t}\right)

@@

For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = 0$ and  $\lim_{t\to\infty} D(t) =  N_0\frac{\mu}{\mu+\lambda} $.
It follows immediately that $\lim_{t\to\infty} G(t) = N_t-N_0\frac{\mu}{\mu+\lambda} $. The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.

@@

    N_0&= 2\int_0^{t_L} \lambda L(t)e^{\alpha(t_L-t)}dt\\

@@

    G &\stackrel{\gamma(t)}{\longrightarrow} & L

@@

        \dot{D} &= \mu  L 

@@

In contrast to the other results, here the final values of species $A$ and $B$ have larger/smaller counts respectively.
This illustrates that spatial effects can shift the final results of growth curves towards both directions, leading to an overall increase/decrease of one species.

@@

        \subcaption{Rand. $D=30.0$\unit{\micro\metre\squared\per\minute}}

@@

        \subcaption{Random initial cell distribution, $D_R=D_I=30.0$\unit{\micro\metre\squared\per\minute}}
